
---
title: "Map Visualizations"
output:
  html_document: default
  fig_width: 10 
  fig_height: 10
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Visualization through maps in R

##### Data Viz for Public Policy
##### October 29th, 2017
#
This is Assignment 5 of the class "Data Visualization in Public Policy" at UChicago (Fall, 2017) and involves producing meaningful maps and narratives to accompany them. 

```{r loading data, echo = FALSE, message=FALSE, warning=FALSE}
library("dplyr")
library("readr")
library("ggplot2")
library("lubridate")
library("knitr")
library("ggthemes")
library("tidyverse")
library("stringr")
library("rgdal")
library("mapproj")
library("scales")
library("gpclib")
library("sp")
library("rgeos")
library("rgdal")
library("maptools")

theme_simple_no_border <- theme(plot.title = element_text(family="Palatino", face="bold", size=18, hjust=0, margin = margin(0, 22, 0, 0)),
                                plot.subtitle=element_text(size=13, family="Georgia", hjust=0, face="italic", color="#8E8883", margin = margin(10, 22, 22, 0)),
                                axis.title = element_text(family = "Franklin Gothic Book",size=12, color="#000000", margin = margin(22, 0, 22, 0)),
                                axis.text=element_text(size=10),
                                plot.caption=element_text(family = "Georgia", hjust = 0, size=9, face="italic", color="black")
)

theme_map <- theme(axis.ticks = element_blank(),
                   axis.text.x = element_blank(),
                   axis.text.y = element_blank(),
                   axis.title.x=element_blank(),
                   axis.title.y=element_blank(),
                   axis.line=element_blank(),
                   panel.background = element_blank(),
                   panel.border = element_blank(), 
                   plot.title = element_text(family="Palatino", face="bold", size=18, hjust=0, margin = margin(0, 22, 0, 0)), 
                   plot.subtitle=element_text(size=13, family="Georgia", hjust=0, face="italic", color="#8E8883", margin = margin(10, 22, 22, 0)), 
                   plot.caption=element_text(family = "Georgia", hjust = 0, size=9, face="italic", color="black")
)

theme_map_2 <- theme(axis.ticks = element_blank(),
                   axis.text.x = element_blank(),
                   axis.text.y = element_blank(),
                   axis.title.x=element_blank(),
                   axis.title.y=element_blank(),
                   axis.line=element_blank(),
                   panel.border = element_blank(),
                   plot.title = element_text(family="Palatino", face="bold", size=18, hjust=0, margin = margin(0, 22, 0, 0)), 
                   plot.subtitle=element_text(size=13, family="Georgia", hjust=0, face="italic", color="#8E8883", margin = margin(10, 22, 22, 0)), 
                   plot.caption=element_text(family = "Georgia", hjust = 0, size=9, face="italic", color="black")
)

```


```{r map1, echo = FALSE, fig.width=9, fig.height=7}

# MAP 1:

## read the shapefile with ReadOGR:

chicago_beat_map <- readOGR(dsn="ChicagoPoliceBeats", layer="ChicagoPoliceBeats")

chicago_beat.points <- fortify(chicago_beat_map, region="beat_num")

# read in the dataset on complaints against police officers (by beats) obtained from Invisible Institute website: https://github.com/invinst/chicago-police-data
chicago_police_complaints <- read.csv("complaints.csv")

# filter only the data for year 2016
chicago_police_complaints$Year <- format(as.Date(chicago_police_complaints$incident_datetime),"%Y")
chicago_police_complaints_2016 <- chicago_police_complaints %>% filter(Year == "2016")

# use string pads to ensure consistency in "beat" codes in both shape files and complaints dataset

chicago_police_complaints_2016$beat <- str_pad(chicago_police_complaints_2016$beat, 4, side="left", pad = 0)

# perform descriptive statistics by beat

count_complaints_per_beat <- chicago_police_complaints_2016 %>% group_by(beat) %>% summarise(freq = n())

# Combine the dataset on complaints against police (by beats) with the beats points
chicago_beat_2016 <- left_join(chicago_beat.points, count_complaints_per_beat, by = c("id" = "beat"))

ggplot(data=chicago_beat_2016, aes(long, lat, group = group, fill = freq)) +
  geom_polygon() +
  geom_path(color="#E5E4D4") +
  coord_equal(1.2) +
  theme_map + 
  scale_fill_distiller(name = "No. of complaints \nagainst police officers", palette = "Spectral", limits = c(0, 30), breaks=seq(0,30,by=10)) + 
  labs(title = "Complaints against police highest in Chicago's South Side", subtitle="South Side and Far Southest Side generated highest number of complaints \nagainst police offciers in 2016.", caption = "Source: Invisible Institute and City of Chicago")
```

Highest number of complaints tend to generate against police officers who are in beats in South Side and Far Southeast Side Chicago. The number of complaints against police officers decrease as one travels from the South to the North of Chicago.


```{r map2, echo = FALSE, fig.width=10, fig.height=8}
## read in Pakistan's districts shape file
pakistan_district_map <- readOGR(dsn="pakistan_district", layer="pakistan_district")

pak_district.points <- fortify(pakistan_district_map, region="district")

## read in data on Pakistan Annual Status of Education Report
## Obtained from: Pakistan Data Portal (http://data.org.pk/frontend/web/masterdatasetdetails/index?dataset_id=433)

pak_educ_status <- read.csv("Annual_status_of_education_report_Pakistan.csv") 

pak_educ_status$District <- as.character(pak_educ_status$District) 
pak_educ_status$District <-  tolower(pak_educ_status$District) 

pak_educ_status <- pak_educ_status %>% filter(Indicators == "Dropouts" & Age == "6-16 years")
pak_educ_status$Value <- (as.numeric(pak_educ_status$Value))/100

pak_district_educ.df <- left_join(pak_district.points, pak_educ_status, by = c("id" = "District"))

pak_cities <- data.frame(
  city= c("Islamabad","Lahore","Quetta","Peshawar","Karachi"),
  long=c(73.066667,  74.343611, 67.000000, 71.583333,67.01),
  lat=c(33.716667, 31.549722, 30.183333, 34.016667,24.86),
  hjust=c(0,0,1,1,1),
  vjust=c(-0.5,-0.5,-0.5,-0.5,0)
)

pak_cities_format <- list( xlab(""), ylab(""), coord_map("polyconic"), geom_point(aes(long,lat),data=pak_cities,size=2), geom_text(aes(x=long,y=lat,label=city,hjust=hjust,vjust=vjust),face="bold",size=3,data=pak_cities)
)

ggplot(data=pak_district_educ.df, aes(long, lat, group = group, fill = Value)) +
  geom_polygon() +
  geom_path(color="#E5E4D4") +
  coord_equal(1) +
  theme(legend.position="bottom", legend.justification = c(0, 1), legend.text=element_text(size=7), legend.background = element_rect(colour = NA), legend.key = element_rect(colour = "white", fill = NA)) +
  theme_map + 
  #pak_cities_format +
  scale_fill_distiller(name = "School dropout percentage", palette = "RdBu", limits = c(0, 0.25), breaks=seq(0,0.25,by=0.05), labels = percent) +
  labs(title = "School drop out rate high in Southern Punjab and Eastern Balochistan", subtitle="High percentage of students aged between 6 and 16 years dropped out \nfrom school in 2014 in Balochistan and South of Punjab", caption = "Source: Idara-E-Taleem-O-Aagahi (ITA)")
  
```

While the 2014 data on drop out rate of students aged between 6 and 16 years for some districts is not available, it can be seen that the highest school drop out rate is registered in Southern Punjab and Mid- and Southern Balochistan. The school drop out rate in the North of Pakistan is relatively better than the average.

```{r map3, echo = FALSE, fig.width=10, fig.height=8}
### map of 2014 exependiture spent on education in Pakistan

# read in the exepnditure data (source: http://data.org.pk/frontend/web/masterdatasetdetails/index?dataset_id=70)
pak_exp_2014 <- read.csv("85-PRSP 2013-14_Pakistan_Educ_expenditure.csv")
pak_exp_2014 <- pak_exp_2014 %>% filter(Description == "Total primary education expenditure (in million)" | Description == "Total secondary education expenditure (in million)")

pak_exp_2014$Value <- pak_exp_2014$Value/105.4 # (current exchange rate: USD 1= PKR 105.4)

count_educspend_per_province <- pak_exp_2014 %>% group_by(Province) %>% summarise(educ_sum = sum(Value))
count_educspend_per_province$Province <- as.character(count_educspend_per_province$Province)
count_educspend_per_province$Province[[3]] = "Islamabad"
count_educspend_per_province$Province[[4]] = "KP"
count_educspend_per_province$Province[[5]] = "PUNJAB"
count_educspend_per_province$Province[[6]] = "SINDH"

#get the province points from shape file
pak_provinces.points <- fortify(pakistan_district_map, region="province")

# join data frames
pak_educexp_educ.df <- left_join(pak_provinces.points, count_educspend_per_province, by = c("id" = "Province"))

ggplot(data=pak_educexp_educ.df, aes(long, lat, group = group, fill = educ_sum)) +
  geom_polygon() +
  geom_path(color="#E5E4D4") +
  coord_equal() +
  theme(legend.position="bottom", legend.justification = c(0, 1), legend.text=element_text(size=8), legend.background = element_rect(colour = NA), legend.key = element_rect(colour = "white", fill = NA)) +
  theme_map  +
  scale_fill_distiller(name = "Education Expenditure \n(USD million)", limits = c(0, 450), breaks=seq(0,450,by=100)) +
  labs(title = "Education expenditure in Punjab is ten times that of Balochistan", subtitle="Education Expenditure in Punjab was highest whereas lowest in the \nlargest province of Pakistan in 2014.", caption = "Source: Pakistan Ministry of Finance")
```

The previous figure showed that drop out rate is highest in Southern Punjab and Southern Balochistan. This picture shows that the education expenditure in 2014 was highest in the province of Punjab, the most industrial and economically active province of Pakistan. It is possible that most education expenditure in Punjab was allocated to the northern or middle areas of the province.

Balochistan, the largest province of Pakistan, reported the lowest amount of education expenditure -- almost the same amount as that of Islamabad, the capital city of Pakistan. This explains the reason why many parts of Balochistan had above average school drop out rate.